---
title: "Snowshoe Hares"
author: "Patrick Pelegri-O'Day"
date: "11/5/2021"
output: html_document
---

## Introduction
A 4 - 5 sentence "mini-introduction" to what this study is exploring, including basic background information about the subject, site and study (with citations as needed), and a brief overview of what is included in the exploratory report. 

## Data and Analyses
A 3 - 4 sentence section in which you briefly describe the data, including citation & credit as needed, and provide an overview of the main analysis, statistical methods and tools used.

**Data citation:** Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14

**Link to metadata:** https://portal.edirepository.org/nis/metadataviewer?packageid=knb-lter-bnz.55.22

## Setup

Attach packages
```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(kableExtra)
library(lubridate)

library(effsize)
library(broom)
library(ggbeeswarm)


```

Read in the data
```{r}
hares_raw <-
  read_csv(show_col_types = FALSE, here("data", "bonanza_hares.csv"))
```


## Data exploration

Convert date column into YYYY-MM-DD format, change class to Date, create column year
```{r}
hares <- hares_raw %>% 
  mutate(date = mdy(date)) %>% # replace original date column with new date column in YYYY-MM-DD format in which column class is "Date"
  mutate(year = year(date)) # create new column called year that contains the year, derived from the date column
```

Annual juvenile hare trap counts. Count the total number of juvenile hare trappings during each year of the study, and create a finalized data visualization of the counts by year. Include a figure caption below the figure in your report. 
```{r}
hares_trap_counts <- hares %>% 
  group_by(year) %>% 
  summarize(count = n())

kable(hares_trap_counts, # create table
      col = c("Year", "Count"), # assign column names
      caption = "Table 1: Count of Trapped Juvenile Hares") %>% 
  kable_styling(full_width = FALSE) # make table narrower
```

