---
title: "Snowshoe Hares"
author: "Patrick Pelegri-O'Day"
date: "11/5/2021"
output: html_document
---

## Overview

#### Introduction **UPDATE** **Consider subheaders throughout document**
A 4 - 5 sentence "mini-introduction" to what this study is exploring, including basic background information about the subject, site and study (with citations as needed), and a brief overview of what is included in the exploratory report. 

#### Data and Analyses **UPDATE**
A 3 - 4 sentence section in which you briefly describe the data, including citation & credit as needed, and provide an overview of the main analysis, statistical methods and tools used.

## Setup

#### Attach packages and read in data

Attach packages
```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(kableExtra)
library(lubridate)
library(effsize)
library(broom)
library(ggbeeswarm)


```

Read in the data
```{r}
hares_raw <-
  read_csv(show_col_types = FALSE, here("data", "bonanza_hares.csv"))
```

#### Initial data wrangling

Filter to retain only juvenile hares, convert date column into YYYY-MM-DD format, change class to Date, create column year, create column grid_full with full site names, replace values in column sex with fully spelled out labels
```{r}
hares <- hares_raw %>% 
  filter(age == "j") %>% # filter to only retain observations for juvenile hares
  mutate(date = mdy(date)) %>% # replace original date column with new date column in YYYY-MM-DD format in which column class is "Date"
  mutate(year = year(date)) %>% # create new column called year that contains the year, derived from the date column
  mutate(grid_full = case_when( # add a new column grid_full, which contains full site names taken from metadata
    grid %in% c("bonrip") ~ "Bonanza Riparian",
    grid %in% c("bonmat") ~ "Bonanza Mature",
    grid %in% c("bonbs") ~ "Bonanza Black Spruce")) %>% 
  mutate(sex = case_when( # change variable names for sex
    sex %in% c("m") ~ "Male",
    sex %in% c("f") ~ "Female",
    sex %in% c("NA") ~ "NA"))
```

## Data analysis and visualization

#### I) Juvenile snowshoe hare trap counts

Visualize the total number of juvenile hare trappings counted during each year of the study
```{r, fig.cap = "**Figure 1: Count of Juvenile Snowshoe Hares.** This graph displays counts of juvenile snowshe hares that were trapped in Bonanza Creek Experimental Forest. The observation period spans 1998 to 2012."}
# Group hares by year and create count per year
hares_trap_counts <- hares %>% 
  group_by(year) %>% 
  summarize(count = n())

# Visualize count of trapped hares by year
ggplot(hares_trap_counts,
       aes(x = year, y = count)) +
  geom_col(fill = "deepskyblue4",
           width = 0.6) +
  labs(x = "\nYear", y = "Count\n") + # increase margins between axes labels and plot
  theme(axis.title = element_text(face = "bold",
                                  size = 11)) +
  theme_minimal()
```

Create data frame containing maximum, minimum, and median annual counts of juvenile snowshoe hares
```{r}
# Group hares by year and calculate count per year as well as minimum count, maximum count, and median count
hares_counts_stats <- hares_trap_counts %>% # start from hare counts df
  summarize(min_count = min(count), # calculate minimum, maximum, and median count for reference in-line in text below
            max_count = max(count),
            med_count = median(count))
```

The minimum count of juvenile snowshoe hares trappings was **`r hares_counts_stats$min_count`**, and the maximum count was **`r hares_counts_stats$max_count`**. Note that there is no data for years 2003 and 2009. The median count of juvenile snowshoe hares trappings, rounded to the nearest integer, was **`r round(hares_counts_stats$med_count,0)`**. We see a general decline in the number of juvenile snowshoe hares trap counts during the study period. Within that general trend, there is significant variation in trap counts from year to year. The annual variation in counts may reflect variation in hare population, or may reflect bias in the samples due to inconsistent sampling methods. To standardize data collection on snowshoe hare trap counts, I recommend conducting future counts by conducting a consistent series of transects at the same time of year.

#### II) Juvenile hare weights

Create a faceted plot comparing juvenile hare weights by sex and site, including groups where these variables are not recorded (NA)
```{r, fig.cap = "**Figure 2: Juvenile hare weights by sex and site.** These three graphs display juvenile hare weights (g) by sex. The graphs are faceted by site of observation."}

ggplot(data = hares, 
       aes(x = sex, y = weight)) +
  geom_beeswarm(aes(color = grid_full)) + # use geom_beeswarm to complement box plot, use color to designate site
  scale_color_manual(values = c("chocolate2","darkolivegreen4","plum3")) +
  geom_boxplot(fill = NA, width = 0.2, outlier.color = NA) + # apply box plots to show quartiles and median, make them transparent, don't highlight outliers
  theme_minimal() +
  facet_wrap(~grid_full) + # facet graph by site
  labs(x = "\nSex", # increase margin between x label and plot
       y = "Weight (g)\n", # increase margin between y label and plot
       color = "Site")
```


This figure shows juvenile snowshoe hare weights by sex at three different sites. Box plots show the quartile spread of observations. Median hare weight for each category is shown as a black bar. This graph shows that median male juvenile hare weight is higher than median female juvenile hare weight. The spread of weights observed is quite large, particularly at the Bonanza Riparian site, which has the greatest number of observations of the three sites.

#### III) Male and female juvnile hare weight comparison

Compare mean and standard deviation of hare weight by sex and site
```{r}
# Create a data frame from hares that groups by sex  and calculates summary statistics for hare weight
hares_wt_sex <- hares %>% 
  group_by(sex) %>% 
   summarize(mean = round(mean(weight, na.rm = TRUE), 1),
             sd = round(sd(weight, na.rm = TRUE), 1),
             sample_size = n())


# Create a series of tables that display the summary statistics calculated in the data frames above
kable(hares_wt_sex,
      col = c("Sex", "Mean Weight (g)", "Standard deviation", "Sample Size"),
      caption = "Table 2: Summary statistics on weight of juvenile hares by sex") %>%
  kable_styling(full_width = FALSE)
```

```{r, include = FALSE}
# Data exploration to determine whether t.test is appropriate

# Pull vectors of weights for male and female juvenile snowshoe hares

f_hares <- hares %>% 
  filter(sex == "Female") %>% 
  pull(weight)

m_hares <- hares %>% 
  filter(sex == "Male") %>% 
  pull(weight)

# plot histograms & qq-plots to check assumptions
hist(f_hares) # Looks normal
qqnorm(f_hares) # Looks linear
hist(m_hares) # Looks pretty normal
qqnorm(m_hares) # Looks relatively linear

# Overall: assumptions of normality hold. Because of large sample size (>1000) Central Limit Theorem holds and we expect means to be normally distributed. So, two-sample t-test is ok for means comparison.
```

Compare means: difference in means, effect size, unpaired two-sample t-test
```{r}
# create vector m_hares that contains weights only of female hares
m_hares <- hares %>% 
  filter(sex == "Male") %>% 
  pull(weight)

m_hares <- m_hares[!is.na(m_hares)] # remove NA values from f_hares

# create vector m_hares that contains weights only of male hares
f_hares <- hares %>% 
  filter(sex == "Female") %>% 
  pull(weight)

f_hares <- f_hares[!is.na(f_hares)] # remove NA values from m_hares

# run t-test
hares_t <- t.test(m_hares, f_hares)

# get tidy model results to call outputs in-line: 
hares_t_tidy <- tidy(hares_t)

# get effect size using cohen.d()
hares_cohen <- cohen.d(m_hares, f_hares)

# find means and SDs for weights of male & female hares
male_mean <- mean(m_hares)
female_mean <- mean(f_hares)
male_sd <- sd(m_hares)
female_sd <- sd(f_hares)
```

On average, juvenile male snowshoe hares are heavier than juvenile female snowshoe hares (`r round(male_mean,2)` $\pm$ `r round(male_sd, 2)` g and `r round(female_mean,2)` $\pm$ `r round(female_sd, 2)` g, respectively. The absolute difference in mean weights is `r round(male_mean, 2) - round(female_mean, 2)` g, which is a difference of `r round((male_mean - female_mean)/((male_mean + female_mean)/2)*100, 2)`%. This difference in means is significant (Welch's two-sample t-test: t(`r round(hares_t_tidy$parameter,2)`) = `r round(hares_t_tidy$statistic,2)`, p-value = `r hares_t_tidy$p.value`, and the effect size is small to moderate (Cohen's *d* = `r round(hares_cohen$estimate,2)`).

#### IV) Relationship between hare weight and hind foot length

Explore linear regression relating juvenile snowshoe hare weight and hind foot length
```{r}
# run linear regression
hares_lm <- lm(weight ~ hindft, data = hares)

# get tidy versions of the model output to call later on in text
hares_lm_tidy <- tidy(hares_lm)
hares_lm_glance <- glance(hares_lm)

# get Pearson's r correlation
hares_cor <- cor.test(hares$weight, hares$hindft)

# get tidy version of correlation output
hares_cor_tidy <- tidy(hares_cor)
```

```{r, include = FALSE}
# visualize model diagnostics
plot(hares_lm)
```


Plot linear regression
```{r}

ggplot(data = hares, 
       aes(x = hindft, y = weight)) +
  geom_point(color = "deepskyblue4") +
  geom_smooth(method = "lm", se = FALSE, color = "seashell3") +
  theme_minimal() +
  labs(x = "\nHind foot length (mm)",
       y = "Weight (g)\n")
```
<br />**Figure 3: Correlation of hare weight and hind foot weight.** This figure shows a scatter plot of juvenile snowshoe hare weight (g) as a function of hind foot length (mm). Points indicate individual hare measurements. A line of best fit derived by ordinary least squares is superimposed on the scatter plot. Linear model summary: $\beta$~1~ = `r round(hares_lm_tidy$estimate[2], 2)` g mm^-1^, p < 0.001, R^2^ = `r round(hares_lm_glance$r.squared,2)`, Pearson's *r* = `r round(hares_cor_tidy$estimate,2)`). Data: Kielland et al. (2017).

The relationship between juvenile snowshoe hare weight and hind foot length appears linear but with significant heteroscedasticity (Figure 3). Single variable linear regression showed that hind foot length (mm) significantly predicts juvenile hare weight (p < 0.001, R^2^ = `r round(hares_lm_glance$r.squared,2)`) with an average slope of $\beta$ = `r round(hares_lm_tidy$estimate[2], 2)` g mm^-1^ (i.e., for each one millimeter increase in hind foot length we expect an average increase in weight of `r round(hares_lm_tidy$estimate[2], 2)` g). Juvenile hare hind foot length and weight have a medium positive correlation that is statistically significant (Pearson's *r* = `r round(hares_cor_tidy$estimate,2)`, p < 0.001). Diagnostic plots, not included here, display residuals that are heteroscedastic and are somewhat normally distributed. The assumption of homoscedasticity does not appear to be satisfied, so linear regression is not a suitable tool for analyzing these data in their current form.

## Summary **CHANGE FOR HARES**

Exploratory data analysis reveals the following initial findings: 

- Male juvenile snowshoe hares have longer mean hind foot length than female juvenile snowshoe hares (the difference is significant, and the effect size is large)
- Flipper length appears linearly related with body mass across all three penguin species; the relationship (by simple linear regression) is significant, with a slope (expected average increase in body mass with a 1 mm increase in flipper length) of `r round(hares_lm_tidy$estimate[2], 2)` g mm^-1^.

## Citations 

**Data citation:** Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14

**Link to metadata:** https://portal.edirepository.org/nis/metadataviewer?packageid=knb-lter-bnz.55.22